---
title: "Bayesian stats assignments 1"
author: "Edwin and Rick"
date: "29/11/2017"
output: html_document
---

This set of assignments accompanies the first slides. It comprises three parts, about probabilities, conjugate priors and acceptance-rejection sampling. 

If you'd rather complete these assignments in python than R, than you are most welcome to do so.

## 1 Conjugacy

The beta distribution is a natural prior to the parameter of the binomial distribution, because of the conjugacy. 

1.1 Write a function that takes alpha and beta as its parameters and plots the density of the beta distribution with the given parameters on a fine grid between 0 and 1.

```{r, message = FALSE}
library(tidyverse)
plot_beta <- function(alpha, beta) {
  plot_data   <- data.frame(x = seq(0, 1, by = .005))
  plot_data$y <- dbeta(plot_data$x, alpha, beta)
  ggplot(plot_data, aes(x, y)) + geom_line()
}
```

1.2 Play around with the parameters to get a feel for the beta distribution. Can you find parameter settings that give a uniform prior (equal credibility over between 0 and 1)?

```{r}
plot_beta(1, 1)
```

1.3 Arjan's parents have 7 grand sons and no grand daughters, is this chance or is there a boy predisposition for this generation Van den Heuvels? Let $x$ be the probability that a baby in this generation of the family is a boy. Use the uniform prior from 2.2, what is the posterior?

```{r}
plot_beta(8, 1)
```

1.4 Interpret the distribution.

```{r}
1 - pbeta(0.5, 8, 1)

# almost all credibility right of 0.5, seems like a strong tendency towards boys in the family.
```

1.5 Do we have an a priori belief about the probability on getting a boy or girl? Is a uniform prior realistic? Redo the analysis with a more informed prior.

```{r}
# it is well known that the probability of having a boy is around 0.5 and there
# is no research indicating dependency between first and later born.

# a prior like 20, 20 seems justified
plot_beta(20, 20)
plot_beta(27, 20)
```

With this prior we see it is very well likely due to chance that there are 7 VDH boys.

## 2 Acceptance-rejection

The nice and juicy website moordatlas.nl records all the cases of murder for the last year. The extract moordata.csv contains the weekly counts of murders on this website. 

2.1 Read this data.

```{r, message = FALSE}
moord <- read_csv("moorddata.csv")
```

2.2  Lets assume that the number of murders in one week is independent of the number of murders in the weeks before. What would be a suitable distribution to model this data?

https://en.wikipedia.org/wiki/Poisson_distribution

2.3 What is the expected value of this distribution? Lets say we expect 1 or 2 murders most weeks, what value of lambda would be suitable?

```{r}
data_frame(x = 0:10, y = dpois(x, 2)) %>% 
  ggplot(aes(x, y)) + geom_bar(stat = "identity", width = .5)
```

2.4 Could we use a Normal(2, 1) prior for lambda? If not, what would be a suitable distribution?

```{r}
# lambda is strictly positive so we should not use a distribution that has a domain on negative values
prior_df <- data_frame(x = seq(0, 12, .01),
                       y = dgamma(seq(0, 12, .01), shape = 2, scale = 2))
ggplot(prior_df, aes(x, y)) + geom_line()
```


2.5 Plot the data and derive the likelihood for lambda.

```{r}
ggplot(moord, aes(n)) + geom_bar()
```

```{r}
poisson_likelihood <- function(lambda, x) {
  prod(dpois(x, lambda))
}

lik_frame     <- data_frame(lambda = seq(0, 10, by = .05))
lik_frame$lik <- map_dbl(lik_frame$lambda, poisson_likelihood, x = moord$n)
ggplot(lik_frame, aes(lambda, lik)) + geom_line()
```

2.6 Write a function to apply in acceptance-rejection sampling, in which the prior and the likelihood are multiplied for a given value of lambda. 

```{r}
ar_function <- function(lambda,
                        x, 
                        prior      = dgamma(lambda, shape = 2, scale = 2),
                        likelihood = poisson_likelihood(lambda, x)) {
  prior * likelihood
}
```

2.7 Determine the highest value of the prior x likelihood function, which will be your max for the acceptance-rejection sampling. (See peak in the likelihood to determine your grid).

```{r}
lambda_range <- seq(2, 3, by = .001)
max_val <- purrr::map_dbl(lambda_range, ar_function, x = moord$n) %>% max()
```

2.7 Now we do the actual sampling. Sample a large number of observations between 0 and your answer at 2.6 (x1) and (the same) large number on the range 0 and 12 (x2). Accept if the function at 2.5 of x2 is smaller than its corresponding x1 value, reject otherwise. Make a plot as in the slides.

```{r}
n <- 10^5
ar <- data_frame(x1 = runif(n, 0, max_val),
           x2 = runif(n, 0, 12)) %>% 
  mutate(f_x2 = map_dbl(x2, ar_function, x = moord$n)) %>% 
  mutate(accepted = as.character(f_x2 > x1))
ggplot(ar, aes(x2, x1)) + geom_point(aes(col = accepted))
```

2.8 Use the accepted samples to obtain the statistics of the posterior. Get the statiscs and plot the posterior.

```{r}
ar %>% 
  filter(accepted == TRUE) %>% 
  summarise(mn = mean(x2), sd = sd(x2))
```

2.9 Does the prior or the likelihood dominate the posterior? Would you expect this from the shapes of the prior and the likelihood.



